---
title: "WT chimera: Batch correction"
author: "Pijuan-Sala, Griffiths, Guibentif et al.: A single-cell resolution molecular roadmap from mouse gastrulation to early organogenesis."
output:
  BiocStyle::html_document:
    code_folding: hide
    number_sections: yes
    toc: yes  
---


```{r setup, message=FALSE}
library(Matrix)
library(scran)
library(Rtsne)
library(irlba)
library(cowplot)

source("/nfs/research1/marioni/jonny/chimera-wt/scripts/core_functions.R")
load_data()

nPC = 50


```

In this script, we perform batch correction on our data.


#Batch correction

For batch correction, we employ the scran function `fastMNN`, which performs batch correction in the manner of `mnnCorrect`, but in the PC-space, and much faster. Critically, this is a composition-aware batch-correction strategy that should not be affected by the lack of e.g. extraembryonic tissue in the Tomato+ samples (as the injected stem cells do not contribute to these lineages). We correct within each timepoint only.

##Total correction

```{r mnncorrect-1}

hvgs_E7.5 = getHVGs(scater::normalize(sce[,meta$stage == "E7.5"]))
hvgs_E8.5 = getHVGs(scater::normalize(sce[,meta$stage == "E8.5"]))

#perform batch correction within each genotype, then across the genotypes
correct1 = doBatchCorrect(counts = logcounts(scater::normalize(sce[hvgs_E7.5, meta$stage == "E7.5"])), 
                          timepoints = meta$tomato[meta$stage == "E7.5"], 
                          samples = meta$sample[meta$stage == "E7.5"], 
                          timepoint_order = as.logical(c("TRUE", "FALSE")), 
                          sample_order = 1:4)

```

```{r mnncorrect-2}
#just two samples so no need for our wrapper
correct2 = fastMNN(scater::normalize(sce[, meta$sample == 5]),
                   scater::normalize(sce[, meta$sample == 6]),
                   subset.row = rownames(sce) %in% hvgs_E8.5,
                   cos.norm = FALSE,
                   approximate = TRUE)

```

```{r mnncorrect-join}

corrected = list("E7.5" = correct1,
                 "E8.5" = correct2$corrected)

saveRDS(corrected, file = "/nfs/research1/marioni/jonny/chimera-wt/data/corrected_pcas.rds")



```

A t-SNE visualisation of our cells, pre- and post-correction, is shown in Figure \@ref(fig:compare-all).

```{r compare-all, fig.height = 10, fig.wide = TRUE, fig.cap = "t-SNE of cells before and after correction. Red and coral cells are Tomato+ (injected), black and grey cells are Tomato- (embryo). Coral and grey cells mark the thrid and fourth samples in the E7.5 timepoint."}

base_7.5 = prcomp_irlba(t(logcounts(scater::normalize(sce[hvgs_E7.5, meta$stage == "E7.5"]))), n = nPC)$x
base_8.5 = prcomp_irlba(t(logcounts(scater::normalize(sce[hvgs_E8.5, meta$stage == "E8.5"]))), n = nPC)$x

tsne_pre_7.5 = Rtsne(base_7.5, pca = FALSE)$Y
tsne_post_7.5 = Rtsne(corrected$E7.5, pca = FALSE)$Y

tsne_pre_8.5 = Rtsne(base_8.5, pca = FALSE)$Y
tsne_post_8.5 = Rtsne(corrected$E8.5, pca = FALSE)$Y

ro1 = sample(nrow(base_7.5), nrow(base_7.5))
ro2 = sample(nrow(base_8.5), nrow(base_8.5))

p1 = ggplot(as.data.frame(tsne_pre_7.5)[ro1,], aes(x = V1, y = V2, col = factor(meta$sample[meta$stage == "E7.5"][ro1]))) +
  geom_point(size = 0.4) +
  scale_colour_manual(values = c("2" = "black", "1" = "red", "3" = "coral", "4" = "darkgrey")) +
  theme(legend.position = "none",
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  ggtitle("E7.5, pre-correction")

p2 = ggplot(as.data.frame(tsne_post_7.5)[ro1,], aes(x = V1, y = V2, col = factor(meta$sample[meta$stage == "E7.5"][ro1]))) +
  geom_point(size = 0.4) +
  scale_colour_manual(values = c("2" = "black", "1" = "red", "3" = "coral", "4" = "darkgrey")) +
  theme(legend.position = "none",
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  ggtitle("E7.5, post-correction")

p3 = ggplot(as.data.frame(tsne_pre_8.5)[ro2,], aes(x = V1, y = V2, col = factor(meta$sample[meta$stage == "E8.5"][ro2]))) +
  geom_point(size = 0.4) +
  scale_colour_manual(values = c("6" = "black", "5" = "red")) +
  theme(legend.position = "none",
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  ggtitle("E8.5, pre-correction")

p4 = ggplot(as.data.frame(tsne_post_8.5)[ro2,], aes(x = V1, y = V2, col = factor(meta$sample[meta$stage == "E8.5"][ro2]))) +
  geom_point(size = 0.4) +
  scale_colour_manual(values = c("6" = "black", "5" = "red")) +
  theme(legend.position = "none",
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  ggtitle("E8.5, post-correction")

plot_grid(p1, p2, p3, p4)

```


#Celltype plots

Figure \@ref(fig:celltype) shows the same plots, but coloured by the mapped celltype (see the mapping script for details).

```{r celltype, fig.wide = FALSE, fig.height=12, fig.cap = "t-SNEs, coloured by celltype."}

meta_final = meta[!meta$celltype.mapped %in% c("Doublet", "Stripped"),]
corrected_7.5_final = corrected$E7.5[!(meta$celltype.mapped[meta$stage == "E7.5"] %in% c("Doublet", "Stripped")),]
corrected_8.5_final = corrected$E8.5[!(meta$celltype.mapped[meta$stage == "E8.5"] %in% c("Doublet", "Stripped")),]

tsne_final_7.5 = Rtsne(corrected_7.5_final, pca = FALSE)$Y
tsne_final_8.5 = Rtsne(corrected_8.5_final, pca = FALSE)$Y

ro1 = sample(sum(meta_final$stage == "E7.5"), sum(meta_final$stage == "E7.5"))
ro2 = sample(sum(meta_final$stage == "E8.5"), sum(meta_final$stage == "E8.5"))

plegend = ggplot(as.data.frame(tsne_final_7.5)[ro1,], aes(x = V1, y = V2, col = factor(meta_final$celltype.mapped[meta_final$stage == "E7.5"][ro1], levels = names(celltype_colours), ordered = TRUE))) +
  geom_point() +
  scale_colour_manual(values = celltype_colours, drop = FALSE) +
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank())  +
  guides(col = guide_legend(override.aes = list(size = 5)))

p1 = ggplot(as.data.frame(tsne_final_7.5)[ro1,], aes(x = V1, y = V2, col = factor(meta_final$celltype.mapped[meta_final$stage == "E7.5"][ro1], levels = names(celltype_colours), ordered = TRUE))) +
  geom_point() +
  scale_colour_manual(values = celltype_colours) +
  theme(legend.position = "none",
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  ggtitle("E7.5")

p2 = ggplot(as.data.frame(tsne_final_8.5)[ro2,], aes(x = V1, y = V2, col = factor(meta_final$celltype.mapped[meta_final$stage == "E8.5"][ro2], levels = names(celltype_colours), ordered = TRUE))) +
  geom_point() +
  scale_colour_manual(values = celltype_colours) +
  theme(legend.position = "none",
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  ggtitle("E8.5")

plot_grid(p1, get_legend(plegend), p2, ncol = 1)


```

# UMAP

Finally, we generate UMAP coordinates of the batch-corrected data.

```{r generate-umap}

write.table(corrected_7.5_final, file = "/nfs/research1/marioni/jonny/chimera-wt/scripts/batch_correct/scanpy_input.tab", sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE)

system("python3 /nfs/research1/marioni/jonny/chimera-wt/scripts/batch_correct/umap.py")

umap_7.5 = read.table("/nfs/research1/marioni/jonny/chimera-wt/scripts/batch_correct/scanpy_output.tab", sep = "\t", header = FALSE)


write.table(corrected_8.5_final, file = "/nfs/research1/marioni/jonny/chimera-wt/scripts/batch_correct/scanpy_input.tab", sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE)

system("python3 /nfs/research1/marioni/jonny/chimera-wt/scripts/batch_correct/umap.py")

umap_8.5 = read.table("/nfs/research1/marioni/jonny/chimera-wt/scripts/batch_correct/scanpy_output.tab", sep = "\t", header = FALSE)

```

```{r plot-umap, fig.height = 12}

p1 = ggplot(as.data.frame(umap_7.5)[ro1,], aes(x = V1, y = V2, col = factor(meta_final$celltype.mapped[meta_final$stage == "E7.5"][ro1], levels = names(celltype_colours), ordered = TRUE))) +
  geom_point() +
  scale_colour_manual(values = celltype_colours) +
  theme(legend.position = "none",
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  ggtitle("E7.5")

p2 = ggplot(as.data.frame(umap_8.5)[ro2,], aes(x = V1, y = V2, col = factor(meta_final$celltype.mapped[meta_final$stage == "E8.5"][ro2], levels = names(celltype_colours), ordered = TRUE))) +
  geom_point() +
  scale_colour_manual(values = celltype_colours) +
  theme(legend.position = "none",
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  ggtitle("E8.5")

plot_grid(p1, plegend, p2, ncol = 1)

ggsave(p1, file = "/nfs/research1/marioni/jonny/chimera-wt/scripts/batch_correct/umap_7.5.pdf",
       width = 5, height = 5)
ggsave(p2, file = "/nfs/research1/marioni/jonny/chimera-wt/scripts/batch_correct/umap_8.5.pdf",
       width = 5, height = 5)

saveRDS(list("E7.5" = umap_7.5,
             "E8.5" = umap_8.5),
        file = "/nfs/research1/marioni/jonny/chimera-wt/data/umaps.rds")

```


#Session Info
```{r sessinf}
sessionInfo()
```
