---
title: "Batch correction"
author: "Pijuan-Sala, Griffiths, Guibentif et al.: A single-cell resolution molecular roadmap from mouse gastrulation to early organogenesis."
output:
  BiocStyle::html_document:
    code_folding: hide
    number_sections: yes
    toc: yes  
---


```{r setup, message=FALSE}
library(Matrix)
library(scran)
library(Rtsne)
#set it up for scran to be properly parallelised
library(BiocParallel)
ncores = 16
mcparam = SnowParam(workers = ncores)
register(mcparam)
library(irlba)
library(cowplot)

source("/nfs/research1/marioni/jonny/embryos/scripts/core_functions.R")
load_data(remove_doublets = TRUE, remove_stripped = TRUE)

nPC = 50

# #subset for testing
# sce = normalize(sce[,meta$stage %in% c("E7.25", "E7.5")])
# meta = meta[meta$stage %in% c("E7.25", "E7.5"),]

```

In this script, we perform batch correction on our data.


#Batch correction

For batch correction, we employ the scran function `fastMNN`, which performs batch correction in the manner of `mnnCorrect`, but in the PC-space, and much faster. Specifically, we will perform 3 types of correction:

1. Correction of entire dataset, using PCs calculated over all cells using HVGs from all cells

2. Correction per timepoint, using PCs calculated for cells of each timepoint, using HVGs calculated at each timepoint

3. Correction per Theiler stage, using PCs calculated for cells of each Theiler stage, using HVGs calculated at each Theiler stage

Three types of correction are important: considering each timepoint/Theiler stage separately will provide a greater resolution of the structure of the data rather than subsetting coordinates from a single large PCA.

##Total correction

```{r mnncorrect-all}

hvgs = getHVGs(sce)

#get order: oldest to youngest; most cells to least cells
order_df = meta[!duplicated(meta$sample), c("stage", "sample")]
order_df$ncells = sapply(order_df$sample, function(x) sum(meta$sample == x))
order_df$stage = factor(order_df$stage, 
                        levels = rev(c("E8.5", 
                                   "E8.25", 
                                   "E8.0", 
                                   "E7.75", 
                                   "E7.5", 
                                   "E7.25", 
                                   "mixed_gastrulation", 
                                   "E7.0", 
                                   "E6.75", 
                                   "E6.5")))
order_df = order_df[order(order_df$stage, order_df$ncells, decreasing = TRUE),]
order_df$stage = as.character(order_df$stage)

all_correct = doBatchCorrect(counts = logcounts(sce)[rownames(sce) %in% hvgs,], 
                             timepoints = meta$stage, 
                             samples = meta$sample, 
                             timepoint_order = order_df$stage, 
                             sample_order = order_df$sample, 
                             npc = 50,
                             BPPARAM = mcparam)


save(all_correct, file = "/nfs/research1/marioni/jonny/embryos/scripts/batch_correct/all_correction.RData")

```

A t-SNE visualisation of all cells, pre- and post-correction, is shown in Figure \@ref(fig:compare-all).

```{r compare-all, fig.height = 14, fig.wide = TRUE, fig.cap = "Correction over all data."}

base_pca = prcomp_irlba(t(logcounts(sce)[rownames(sce) %in% hvgs,]), n = nPC)$x
base_tsne = Rtsne(base_pca, pca = FALSE)$Y
corrected_tsne = Rtsne(all_correct, pca = FALSE)$Y

base_tsne = as.data.frame(base_tsne)
base_tsne$sample = meta$sample
base_tsne$state = "Uncorrected"

corrected_tsne = as.data.frame(corrected_tsne)
corrected_tsne$sample = meta$sample
corrected_tsne$state = "Corrected"

bc_tsne = rbind(base_tsne, corrected_tsne)

reorder = sample(nrow(bc_tsne), nrow(bc_tsne))

ggplot(bc_tsne[reorder,], aes(x = V1, y = V2, col = factor(sample))) +
  geom_point(size = 0.4) +
  scale_colour_Publication() +
  theme(legend.position = "none",
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  facet_wrap(~state, nrow = 2)

```

##Timepoint correction


```{r mnncorrect-timepoint}

stage_corrections = lapply(unique(meta$stage), function(x){
  sub_sce = normalize(sce[,meta$stage == x])
  sub_meta = meta[meta$stage == x,]
  hvgs = getHVGs(sub_sce)
  
  correct = doBatchCorrect(counts = logcounts(sub_sce)[rownames(sub_sce) %in% hvgs,], 
                           timepoints = sub_meta$stage, 
                           samples = sub_meta$sample, 
                           timepoint_order = order_df$stage, 
                           sample_order = order_df$sample, 
                           npc = 50,
                           BPPARAM = mcparam)
  return(correct)
  
})
names(stage_corrections) = unique(meta$stage)

```


t-SNE visualisations of cells at each timepoint, pre- and post-correction, are shown in Figure \@ref(fig:compare-stage). Corrections were computed in the scope of each timepoint exclusively (i.e. for calculation of HVGs, PCA).

```{r compare-stage, fig.height = 20, fig.wide = FALSE, fig.cap = "Correction computed for each timepoint separately."}

tsnes = lapply(stage_corrections, Rtsne, pca = FALSE)
tsnes = lapply(tsnes, function(x) as.data.frame(x$Y))
for(i in 1:length(tsnes)){
  tsnes[[i]]$stage = unique(meta$stage)[i]
  tsnes[[i]]$sample = meta$sample[meta$stage == unique(meta$stage)[i]]
  tsnes[[i]]$state = "Corrected"
}

base_pcas = lapply(unique(meta$stage), function(x){
  sub_sce = normalize(sce[,meta$stage == x])
  sub_meta = meta[meta$stage == x,]
  hvgs = getHVGs(sub_sce)
  
  pca = prcomp_irlba(t(logcounts(sub_sce)[rownames(sub_sce) %in% hvgs,]), n = nPC)
  return(pca$x)
})
base_tsnes = lapply(base_pcas, Rtsne, pca = FALSE)
base_tsnes = lapply(base_tsnes, function(x) as.data.frame(x$Y))
for(i in 1:length(base_tsnes)){
  base_tsnes[[i]]$stage = unique(meta$stage)[i]
  base_tsnes[[i]]$sample = meta$sample[meta$stage == unique(meta$stage)[i]]
  base_tsnes[[i]]$state = "Uncorrected"
}

big_df = rbind(
  do.call(rbind, base_tsnes),
  do.call(rbind, tsnes)
)

reorder = sample(nrow(big_df), nrow(big_df))

ggplot(big_df[reorder,], aes(x = V1, y = V2, col = factor(sample))) +
  geom_point(size = 0.4) +
  scale_colour_Publication() +
  theme(legend.position = "none",
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  facet_grid(stage~state)


```

##Theiler stage correction

```{r mnncorrect-theiler}

theiler_corrections = lapply(unique(meta$theiler), function(x){
  sub_sce = normalize(sce[,meta$theiler == x])
  sub_meta = meta[meta$theiler == x,]
  hvgs = getHVGs(sub_sce)
  
  correct = doBatchCorrect(counts = logcounts(sub_sce)[rownames(sub_sce) %in% hvgs,], 
                           timepoints = sub_meta$stage, 
                           samples = sub_meta$sample, 
                           timepoint_order = order_df$stage, 
                           sample_order = order_df$sample, 
                           npc = 50,
                           BPPARAM = mcparam)
  return(correct)
  
})

names(theiler_corrections) = unique(meta$theiler)


```

t-SNE visualisations of cells at each Theiler stage, pre- and post-correction, are shown in Figure \@ref(fig:compare-theiler). Corrections were computed in the scope of each Theiler stage exclusively (i.e. for calculation of HVGs, PCA).

```{r compare-theiler, fig.height = 20, fig.cap = "Correction computed for each Theiler stage."}

tsnes = lapply(theiler_corrections, Rtsne, pca = FALSE)
tsnes = lapply(tsnes, function(x) as.data.frame(x$Y))
for(i in 1:length(tsnes)){
  tsnes[[i]]$theiler = unique(meta$theiler)[i]
  tsnes[[i]]$sample = meta$sample[meta$theiler == unique(meta$theiler)[i]]
  tsnes[[i]]$state = "Corrected"
}

base_pcas = lapply(unique(meta$theiler), function(x){
  sub_sce = normalize(sce[,meta$theiler == x])
  sub_meta = meta[meta$theiler == x,]
  hvgs = getHVGs(sub_sce)
  
  pca = prcomp_irlba(t(logcounts(sub_sce)[rownames(sub_sce) %in% hvgs,]), n = nPC)
  return(pca$x)
})
base_tsnes = lapply(base_pcas, Rtsne, pca = FALSE)
base_tsnes = lapply(base_tsnes, function(x) as.data.frame(x$Y))

for(i in 1:length(base_tsnes)){
  base_tsnes[[i]]$theiler = unique(meta$theiler)[i]
  base_tsnes[[i]]$sample = meta$sample[meta$theiler == unique(meta$theiler)[i]]
  base_tsnes[[i]]$state = "Uncorrected"
}

big_df = rbind(
  do.call(rbind, base_tsnes),
  do.call(rbind, tsnes)
)

reorder = sample(nrow(big_df), nrow(big_df))

ggplot(big_df[reorder,], aes(x = V1, y = V2, col = factor(sample))) +
  geom_point(size = 0.4) +
  scale_colour_Publication() +
  theme(legend.position = "none",
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  facet_grid(theiler~state)


```

```{r save}

corrections = list("all" = all_correct, "theiler" = theiler_corrections, "stage" = stage_corrections)
saveRDS(corrections, file = "/nfs/research1/marioni/jonny/embryos/data/corrected_pcas.rds")

```

#Session Info
```{r sessinf}
sessionInfo()
```
