---
title: "Tal1 chimera: Mapping to the atlas"
author: "Pijuan-Sala, Griffiths, Guibentif et al.: A single-cell resolution molecular roadmap from mouse gastrulation to early organogenesis."
output:
  BiocStyle::html_document:
    code_folding: hide
    number_sections: yes
    toc: yes  
  html_document:
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_float:
      collapsed: true
    theme: united
---
#Introduction

In this script we map cells from the wild-type chimera onto our atlas dataset. We use the same approach as described for the wild-type chimera.

```{r load, message = FALSE}
library(Matrix)
library(scran)
library(scater)
library(knitr)
library(pheatmap)
library(edgeR)

source("/nfs/research1/marioni/jonny/embryos/scripts/core_functions.R")
load_data()

atlas_sce = sce
atlas_meta = meta

atlas_sce_full = atlas_sce
atlas_meta_full = atlas_meta


atlas_meta$celltype[atlas_meta$doublet] = "Doublet"
atlas_meta$celltype[atlas_meta$stripped] = "Stripped"


he_atlas = scater::normalize(atlas_sce[, atlas_meta$celltype== "Haematoendothelial progenitors"])
he_atlas_meta = atlas_meta[atlas_meta$celltype== "Haematoendothelial progenitors",]

#Downsample to 10k cells per sample (or maximum number of cells in sample)
keep = lapply(unique(atlas_meta$stage), function(x){
  if(x == "mixed_gastrulation"){
    return(c())
  } else if(sum(atlas_meta$stage == x) < 10000) {
    return(which(atlas_meta$stage == x))
  } else {
    hits = which(atlas_meta$stage == x)
    return(sample(hits, 10000))
  }
})
keep = do.call(c, keep)

atlas_sce = scater::normalize(atlas_sce[,keep])
atlas_meta = atlas_meta[keep,]

source("/nfs/research1/marioni/jonny/chimera-tal1/scripts/core_functions.R")
load_data()


```


```{r functions}
getmode <- function(v, dist) {
  tab = table(v)
  #if tie, break to shortest distance
  if(sum(tab == max(tab)) > 1){
    tied = names(tab)[tab == max(tab)]
    sub = dist[v %in% tied]
    names(sub) = v[v %in% tied]
    return(names(sub)[which.min(sub)])
  } else {
    return(names(tab)[which.max(tab)])
  }
}

mnnMap = function(atlas_pca, atlas_meta, map_pca, map_meta, k_map = 10){
  require(kmknn)
  require(scran)
  
  correct = fastMNN(atlas_pca, map_pca, pc.input = TRUE)$corrected
  atlas = 1:nrow(atlas_pca)
  correct_atlas = correct[atlas,]
  correct_map = correct[-atlas,]
  
  knns = kmknn::queryKNN(correct_atlas, correct_map, k = k_map, get.index = TRUE, get.distance = FALSE)
  
  #get closest k matching cells
  k.mapped = t(apply(knns$index, 1, function(x) atlas_meta$cell[x]))
  celltypes = t(apply(k.mapped, 1, function(x) atlas_meta$celltype[match(x, atlas_meta$cell)]))
  stages = t(apply(k.mapped, 1, function(x) atlas_meta$stage[match(x, atlas_meta$cell)]))
  celltype.mapped = apply(celltypes, 1, function(x) getmode(x, 1:length(x)))
  stage.mapped = apply(stages, 1, function(x) getmode(x, 1:length(x)))
  
  out = lapply(1:length(celltype.mapped), function(x){
    list(cells.mapped = k.mapped[x,],
         celltype.mapped = celltype.mapped[x],
         stage.mapped = stage.mapped[x])
  })
  
  names(out) = map_meta$cell
  
  return(out)
  
}

#meta MUST have a cell column, celltype column and a stage column, spelled exactly like that
mapWrap = function(atlas_sce, atlas_meta, map_sce, map_meta, k = 10, return.list = FALSE){
  require(scran)
  #prevent duplicate rownames
  colnames(map_sce) = paste0("map_", colnames(map_sce))
  map_meta$cell = paste0("map_", map_meta$cell)
  
  big_sce = scater::normalize(cbind(atlas_sce, map_sce))
  hvgs = getHVGs(big_sce)
  big_pca = prcomp_irlba(t(logcounts(big_sce[hvgs,])), n = 50)$x
  rownames(big_pca) = colnames(big_sce) 
  atlas_pca = big_pca[1:ncol(atlas_sce),]
  map_pca = big_pca[-(1:ncol(atlas_sce)),]
  
  #correct the atlas first
  order_df = atlas_meta[!duplicated(atlas_meta$sample), c("stage", "sample")]
  order_df$ncells = sapply(order_df$sample, function(x) sum(atlas_meta$sample == x))
  order_df$stage = factor(order_df$stage, 
                          levels = rev(c("E8.5", 
                                         "E8.25", 
                                         "E8.0", 
                                         "E7.75", 
                                         "E7.5", 
                                         "E7.25", 
                                         "mixed_gastrulation", 
                                         "E7.0", 
                                         "E6.75", 
                                         "E6.5")))
  order_df = order_df[order(order_df$stage, order_df$ncells, decreasing = TRUE),]
  order_df$stage = as.character(order_df$stage)
  
  set.seed(42)
  atlas_corrected = doBatchCorrect(counts = logcounts(atlas_sce[hvgs,]), 
                                   timepoints = atlas_meta$stage, 
                                   samples = atlas_meta$sample, 
                                   timepoint_order = order_df$stage, 
                                   sample_order = order_df$sample, 
                                   pc_override = atlas_pca)
  
  mapping = mnnMap(atlas_pca = atlas_corrected,
                   atlas_meta = atlas_meta,
                   map_pca = map_pca,
                   map_meta = map_meta)
  if(return.list){
    return(mapping)
  }
  
  cn = substr(names(mapping), 5, nchar(names(mapping))) # remove 
  ct = sapply(mapping, function(x) x$celltype.mapped)
  st = sapply(mapping, function(x) x$stage.mapped)
  closest = sapply(mapping, function(x) x$cells.mapped[1])
  
  out = data.frame(cell = cn, celltype.mapped = ct, stage.mapped = st, closest.cell = closest)
  return(out)
  
}

```


#Chimera mapping

```{r map}
# set.seed(42)
# mappings = lapply(unique(meta$sample), function(x){
#   mapWrap(atlas_sce, atlas_meta, sce[-nrow(sce), meta$sample == x], meta[meta$sample == x,])
# })
# 
# mappings_list = lapply(unique(meta$sample), function(x){
#   mapWrap(atlas_sce, atlas_meta, sce[-nrow(sce), meta$sample == x], meta[meta$sample == x,], return.list = TRUE)
# })
# 
# mappings = do.call(rbind, mappings)
# mappings_list = do.call(rbind, mappings_list)
# 
# saveRDS(mappings, file = "/nfs/research1/marioni/jonny/chimera-tal1/scripts/map/mapping_chimera.rds")
# saveRDS(mappings_list, file = "/nfs/research1/marioni/jonny/chimera-tal1/scripts/map/mapping_chimera_list.rds")
mappings = readRDS("/nfs/research1/marioni/jonny/chimera-tal1/scripts/map/mapping_chimera.rds")

meta$stage.mapped = as.character(mappings$stage.mapped[match(meta$cell, mappings$cell)])
meta$celltype.mapped = as.character(mappings$celltype.mapped[match(meta$cell, mappings$cell)])
meta$closest.cell = as.character(mappings$closest.cell[match(meta$cell, mappings$cell)])

```


##Mapping results

To which stages do the chimera cells map? This is shown in Figure \@ref(fig:stage-mapping).

```{r stage-mapping, fig.wide = TRUE, fig.height=8, fig.cap = "Mapping of cells to atlas timepoints. A: Mapping of cells from all chimeras to timepoints. B: Mapping of cells in the first E8.5 chimera biological replicate, split by tomato status. C: Mapping of cells in the second E8.5 chimera biological replicate, split by tomato status."}

p1 = ggplot(mapping = aes(x = meta$stage.mapped)) +
  geom_bar(stat = "count", position = "dodge") +
  labs(x = "Atlas stage", y = "# cells") +
  theme(axis.text.x = element_text(angle = -30, hjust = 0),
        axis.title.x =  element_blank()) +
  scale_x_discrete(breaks = unique(meta$stage.mapped)[order(unique(meta$stage.mapped))])
  
  

pdf1 = table(meta$stage.mapped[meta$sample %in% c(1,3)], meta$tomato[meta$sample %in% c(1,3)])
pdf1 = as.data.frame(sweep(pdf1, 2, colSums(pdf1), "/"))

p2 = ggplot(data = pdf1, mapping = aes(x = Var1, y = Freq, fill = Var2)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "black"), 
                    labels = c("TRUE" = "Tomato+", "FALSE" = "Tomato-"),
                    name = "") +
  labs(x = "Atlas stage", y = "Fraction of cells") +
  ggtitle("E8.5 chimera 1") +
  theme(axis.text.x = element_text(angle = -30, hjust = 0),
        legend.position = "none",
        axis.title.x = element_blank()) +
  scale_x_discrete(breaks = unique(meta$stage.mapped)[order(unique(meta$stage.mapped))])


pdf2 = table(meta$stage.mapped[meta$sample %in% c(2,4)], meta$tomato[meta$sample %in% c(2,4)])
pdf2 = as.data.frame(sweep(pdf2, 2, colSums(pdf2), "/"))

p3 = ggplot(data = pdf2, mapping = aes(x = factor(Var1, levels = unique(meta$stage.mapped)[order(unique(meta$stage.mapped))]), 
                                       y = Freq, fill = Var2)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "black"), 
                    labels = c("TRUE" = "Tomato+", "FALSE" = "Tomato-"),
                    name = "",
                    drop = FALSE) +
  labs(x = "Atlas stage", y = "Fraction of cells") +
  ggtitle("E8.5 chimera 2") +
  theme(axis.text.x = element_text(angle = -30, hjust = 0),
        legend.position = "none",
        axis.title.x = element_blank()) +
  scale_x_discrete(breaks = unique(meta$stage.mapped)[order(unique(meta$stage.mapped))], 
                   drop = FALSE)

pdf3 = table(meta$stage.mapped, meta$tomato)
pdf3 = as.data.frame(sweep(pdf3, 2, colSums(pdf3), "/"))

p4 = ggplot(data = pdf3, mapping = aes(x = Var1, y = Freq, fill = Var2)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "black"), 
                    labels = c("TRUE" = "Tomato+", "FALSE" = "Tomato-"),
                    name = "") +
  labs(x = "Atlas stage", y = "Fraction of cells") +
  ggtitle("E8.5 chimeras") +
  theme(axis.text.x = element_text(angle = -30, hjust = 0),
        legend.position = "none",
        axis.title.x = element_blank()) +
  scale_x_discrete(breaks = unique(meta$stage.mapped)[order(unique(meta$stage.mapped))])


grid = plot_grid(p1,p2,p3, p4, labels = "AUTO")

grid

save_plot(grid, file = "/nfs/research1/marioni/jonny/chimera-tal1/scripts/map/plots/stage_map.pdf",
          base_width = 10, base_height = 10)

```

Furthermore, to which celltypes do these cells contribute? The fraction of cells of each celltype that derive from each genotype is shown in Figure \@ref(fig:celltype).

```{r celltype, warning = FALSE, fig.wide = TRUE, fig.cap = "Celltypes", fig.height = 10}


pdfall = table(meta$celltype.mapped, meta$tomato)
colnames(pdfall) = c("TomT", "TomF")[match(colnames(pdfall), c("TRUE", "FALSE"))]
pdfall = sweep(pdfall, 2, colSums(pdfall), "/")
pdfall = as.data.frame.matrix(pdfall)


plegend = ggplot(data = pdfall, mapping = aes(x = TomT, y = TomF, col = rownames(pdfall))) +
  geom_abline(slope = 1, intercept = 0) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(celltype_colours, "Doublet" = "black", "Stripped" = "black"), name = "") +
  labs(x = "Fraction of Tomato+ cells", y = "Fraction of Tomato- cells") +
  ggtitle("all cells") +
  scale_x_log10() +
  scale_y_log10() +
  guides(col = guide_legend(ncol = 5, override.aes = list(size = 5)))

p1 = ggplot(data = pdfall, mapping = aes(x = TomT, y = TomF, col = rownames(pdfall))) +
  geom_abline(slope = 1, intercept = 0) +
  geom_point(size = 4) +
  scale_colour_manual(values = c(celltype_colours, "Doublet" = "black", "Stripped" = "black"), name = "") +
  labs(x = "Tomato+ cells", y = "Tomato- cells") +
  ggtitle("All cells") +
  scale_x_log10(breaks = c(0.1, 0.01, 0.001, 0.0001), labels = c("10%", "1%", "0.1%", "0.01%")) +
  scale_y_log10(breaks = c(0.1, 0.01, 0.001, 0.0001), labels = c("10%", "1%", "0.1%", "0.01%")) +
  theme(legend.position = "none")

pdf1 = table(meta$celltype.mapped[meta$sample %in% c(1,3)], meta$tomato[meta$sample %in% c(1,3)])
colnames(pdf1) = c("TomT", "TomF")[match(colnames(pdf1), c("TRUE", "FALSE"))]
pdf1 = sweep(pdf1, 2, colSums(pdf1), "/")
pdf1 = as.data.frame.matrix(pdf1)

p2 = ggplot(data = pdf1, mapping = aes(x = TomT, y = TomF, col = rownames(pdf1))) +
  geom_abline(slope = 1, intercept = 0) +
  geom_point(size = 4) +
  scale_colour_manual(values = c(celltype_colours, "Doublet" = "black", "Stripped" = "black"), name = "") +
  labs(x = "Tomato+ cells", y = "Tomato- cells") +
  ggtitle("E8.5 chimera 1") +
  scale_x_log10(breaks = c(0.1, 0.01, 0.001, 0.0001), labels = c("10%", "1%", "0.1%", "0.01%")) +
  scale_y_log10(breaks = c(0.1, 0.01, 0.001, 0.0001), labels = c("10%", "1%", "0.1%", "0.01%")) +
  theme(legend.position = "none")



pdf2 = table(meta$celltype.mapped[meta$sample %in% c(2,4)], meta$tomato[meta$sample %in% c(2,4)])
colnames(pdf2) = c("TomT", "TomF")[match(colnames(pdf2), c("TRUE", "FALSE"))]
pdf2 = sweep(pdf2, 2, colSums(pdf2), "/")
pdf2 = as.data.frame.matrix(pdf2)

p3 = ggplot(data = pdf2, mapping = aes(x = TomT, y = TomF, col = rownames(pdf2))) +
  geom_abline(slope = 1, intercept = 0) +
  geom_point(size = 4) +
  scale_colour_manual(values = c(celltype_colours, "Doublet" = "black", "Stripped" = "black"), name = "") +
  labs(x = "Tomato+ cells", y = "Tomato- cells") +
  ggtitle("E8.5 chimera 2") +
  scale_x_log10(breaks = c(0.1, 0.01, 0.001, 0.0001), labels = c("10%", "1%", "0.1%", "0.01%")) +
  scale_y_log10(breaks = c(0.1, 0.01, 0.001, 0.0001), labels = c("10%", "1%", "0.1%", "0.01%")) +
  theme(legend.position = "none")



p = plot_grid(plot_grid(p1, p2, p3), get_legend(plegend), rel_heights = c(0.66, 0.33), ncol = 1)

psave = plot_grid(p1 + coord_fixed(xlim = c(0.00003, 0.25), ylim = c(0.00003, 0.25)), 
                  p2 + coord_fixed(xlim = c(0.00003, 0.25), ylim = c(0.00003, 0.25)), p3 + coord_fixed(xlim = c(0.00003, 0.25), ylim = c(0.00003, 0.25)))

p

save_plot(psave, file = "/nfs/research1/marioni/jonny/chimera-tal1/scripts/map/plots/celltype_map.pdf",
          base_width = 10, base_height = 10)

```


## Doublet rates

The fraction of doublets in each of the 10X samples is shown in Figure \@ref(fig:doublets)


```{r doublets, fig.cap = "Number of doublets called in chimera samples"}

pdf = data.frame(ncells = sapply(1:4, function(x) sum(meta$sample == x)),
                 fracdoubs = c(as.numeric(pdf1["Doublet", 2:1]),
                               as.numeric(pdf2["Doublet", 2:1])))

ggplot(pdf, aes(x = ncells, y= fracdoubs)) +
  geom_point() +
  scale_y_continuous(breaks = c(0,0.05,0.1), labels = c("0%", "5%", "10%"), limits = c(0,0.1)) +
  labs(x = "Cells in sample", y = "Fraction doublets")

```



#Save updated metadata

```{r savemeta}

write.table(meta, file = "/nfs/research1/marioni/jonny/chimera-tal1/data/meta.tab", row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")

```


#Session Info
```{r sessinf}
sessionInfo()
```
